---
title: "Porting a C++ hot loop from Rcpp to RZig"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Porting a C++ hot loop from Rcpp to RZig}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rzig)
compiler <- find_zig(required = FALSE)
has_zig <- isTRUE(compiler$supported)
```

RZig is most useful when an R package has a small, computationally expensive
loop that needs native performance but does not otherwise need C++. This
vignette ports an `axpy`-style operation: return `x + scale * y` without
modifying either R input. Every step up to the installation runs in R without
a Zig compiler; the installation and the calls run when a supported compiler is
found.

## Start from the behavior, not the wrapper

An Rcpp implementation might duplicate `x`, validate the inputs, and mutate the
duplicate in place:

```cpp
#include <Rcpp.h>

// [[Rcpp::export]]
Rcpp::NumericVector axpy_rcpp(
    Rcpp::NumericVector x,
    const Rcpp::NumericVector& y,
    double scale
) {
    if (x.size() != y.size()) Rcpp::stop("x and y must have equal lengths");

    Rcpp::NumericVector result = Rcpp::clone(x);
    for (R_xlen_t i = 0; i < result.size(); ++i) {
        if (i % 100000 == 0) Rcpp::checkUserInterrupt();
        result[i] += scale * y[i];
    }
    return result;
}
```

The important contract is not the C++ class structure. It is that `y` is read
only, `x` is duplicated before mutation, unequal lengths become an R error, and
long computations remain interruptible.

## Scaffold the Zig-backed package

Set the package path explicitly, then run `use_rzig()` once. This disposable
example stays inside R's session temporary directory; for a real project,
replace `package_path` with its explicit package root:

```{r scaffold}
package_path <- tempfile("yourpackage-", tmpdir = tempdir())
dir.create(package_path)
writeLines(
  c(
    "Package: yourpackage",
    "Type: Package",
    "Title: A Small RZig Example",
    "Version: 0.0.1",
    "Authors@R: person('Your', 'Name', email = 'you@example.org', role = c('aut', 'cre'))",
    "Description: Demonstrates a native implementation written in Zig.",
    "License: GPL-3",
    "Encoding: UTF-8"
  ),
  file.path(package_path, "DESCRIPTION")
)
scaffold <- use_rzig(package_path)
scaffold
```

`use_rzig()` creates the portable build files, the Zig source tree with the
framework, a starter `src/rzig/src/main.zig`, and the generated R wrapper for
the starter function. The Zig implementation of the port is shipped with RZig
as `inst/examples/axpy.zig`; copying it over the starter file is the only
edit the port needs:

```{r write-zig, comment = ""}
main_path <- file.path(package_path, "src", "rzig", "src", "main.zig")
file.copy(system.file("examples", "axpy.zig", package = "rzig"), main_path,
  overwrite = TRUE)
cat(readLines(main_path), sep = "\n")
```

`rzig.Mut([]f64)` is the direct expression of copy-on-modify semantics. RZig
duplicates and protects `x` before Zig receives writable storage. The `y`
slice borrows R's numeric storage as read-only memory. Because this function
returns `void`, the generated boundary returns the mutated duplicate
automatically.

The function contains no `SEXP`, `PROTECT`, C++ exception, or registration
table. The types and `@export` marker supply enough information for RZig to
generate those parts.

## Inspect what R will see

`scan_exports()` reads the Zig source and lists every exported function with
its R arguments and the R value of each parameter and of the result:

```{r scan}
scan_exports(package_path)
```

## Generate the bindings

`document()` writes the Zig manifest, the R wrapper and the `NAMESPACE`
block. Run it whenever an exported Zig signature or its documentation
changes. The returned object holds the generated texts:

```{r document}
bindings <- document(package_path)
cat(bindings$wrappers)
cat(bindings$namespace)
```

`rzig_status()` checks every step and tells whether the generated bindings
match the current Zig source and whether a supported Zig compiler is
available for the installation:

```{r status}
rzig_status(package_path)
```

## Check the signature before compiling

`rzig_status()` also rejects, before any compilation, the signatures the
boundary cannot convert. A single-precision argument is one such case:

```{r unsupported}
broken_path <- tempfile("brokenpackage-", tmpdir = tempdir())
dir.create(broken_path)
writeLines(c("Package: brokenpackage", "Version: 0.0.1"),
  file.path(broken_path, "DESCRIPTION"))
invisible(use_rzig(broken_path))
broken_main <- file.path(broken_path, "src", "rzig", "src", "main.zig")
cat("/// @export\npub fn half(x: f32) f32 {\n    return x / 2;\n}\n",
  file = broken_main, append = TRUE)
rzig_status(broken_path)
```

## Build and call

`build_zig()` compiles and installs the package from a copy in the session
temporary directory. It runs the checks of `rzig_status()` first, keeps the
complete installation log, and on failure reports the first errors of the Zig
compiler, the C compiler, the linker or the load test with their files and
lines. The build needs a Zig compiler of a supported release series. The
chunks below run when `find_zig()` finds one; this rendered copy of the
vignette `r if (has_zig) "ran" else "skipped"` them.

```{r install, eval = has_zig}
build <- build_zig(package_path)
build
package_name <- "yourpackage"
loadNamespace(package_name, lib.loc = build$library)
axpy <- getExportedValue(package_name, "axpy")
```

The generated R function has the same three visible arguments:

```{r call, eval = has_zig}
x <- c(1, 2, 3)
y <- c(10, 20, 30)
result <- axpy(x, y, 0.5)
x
result
try(axpy(1:2, c(1, 2, 3), 1))
sum(1:10)
```

The native error is an R condition, and the session remains usable. Integer
vectors are rejected for numeric slices and mutable numeric inputs. Convert
deliberately with `as.numeric()` in R rather than relying on an implicit
native allocation.

```{r cleanup, include = FALSE}
if (has_zig) unloadNamespace(package_name)
unlink(c(package_path, broken_path), recursive = TRUE)
if (has_zig) unlink(build$library, recursive = TRUE)
```

## What changes in the port

Rcpp's `NumericVector` combines an R object, indexing, allocation, and mutation
in one C++ class. RZig separates those capabilities. A borrowed `[]const f64`
cannot mutate its input. `rzig.Mut([]f64)` duplicates before exposing writable
storage. `rzig.Ctx` is used only when Zig-owned scratch or result memory is
needed. Errors are explicit `rzig.Error` values, and the outer boundary turns
them into R conditions after Zig cleanup has finished.

Generated packages compile in Zig's `ReleaseSafe` mode. Bounds checks and
integer-overflow checks therefore remain enabled in installed code. A failure
is routed through RZig's panic boundary instead of silently corrupting adjacent
R memory.

## Measure the boundary separately

Benchmark the algorithm and the language boundary as separate questions. For a
hot loop, compare equivalent copy and mutation semantics, use the same compiler
optimization level, warm each implementation, randomize execution order, and
report distributions rather than one elapsed time.

The RZig repository includes `tests/bench.R`, which checks and times equivalent
RZig, handwritten C, and Rcpp round-trip copies at lengths 1, 1,000, and
1,000,000. It records medians and interquartile ranges from randomized batches.
This isolates conversion overhead from the workload being ported.

## Port incrementally

Move one native function at a time and keep the surrounding R interface stable.
Tests should cover empty and large inputs, missing values, wrong R types,
length mismatches, and continued session use after each error. C libraries can
coexist with Zig during a gradual conversion. C++ classes and templates usually
need to be redesigned as Zig structs and explicit data flow rather than
translated line by line.
